Supermassive black hole and host bulge affairs \\ II. The early and late sequence in the $M_{\rm BH} - M_{\rm *,sph}$ diagram

### 2015-05-29 papers/MbhMsph/mm.tex
fri

\begin{abstract}
In our first paper, we performed a detailed (i.e.~bulge, disks, bars, spiral arms, rings, halo, nucleus, depleted core, etc.) 
decomposition of 66 galaxies, with directly measured black hole masses, 
that had been imaged at $3.6\rm~\mu m$ with \emph{Spitzer}.
Our sample is the largest to date and, for the first time, the decompositions were checked to be consistent with the galaxy kinematics.
Here we present correlations between the black hole mass, $M_{\rm BH}$, 
and the host spheroid (and galaxy) luminosity, $L_{\rm sph}$ (and $L_{\rm gal}$), 
and also stellar mass, $M_{\rm *,sph}$.
Most previous studies have used galaxy samples that were overwhelmingly dominated by high-mass, early-type galaxies.
Instead, our sample includes 17 spiral galaxies, half of which have $M_{\rm BH} < 10^7\rm~M_\odot$, 
and allows us to better investigate the poorly studied low-mass end of the $M_{\rm BH} - M_{\rm *,sph}$ correlation.
The bulges of early-type (E + S0) galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{1.04 \pm 0.10}$, 
consistent with a dry-merging formation scenario,
and define a tight \emph{early-type sequence} with intrinsic scatter $\epsilon_{(Y|X)} = 0.43 \pm 0.06\rm~dex$.
At the low-mass end, the bulges of late-type (Sp) galaxies define a much steeper \emph{late-type sequence}, 
with $M_{\rm BH} \propto M_{\rm *,sph}^{2-3}$, 
indicating that gas-rich processes feed the black hole more efficiently than the host bulge as they coevolve. 
We additionally report that: i) S\'ersic galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{1.48 \pm 0.20}$, a less steep sequence than previously reported; 
ii) bulges with S\'ersic index $n_{\rm sph}<2$, argued by some to be pseudo-bulges, 
are not offset to lower $M_{\rm BH}$ from the correlation defined by bulges with $n_{\rm sph}>2$; 
and iii) $L_{\rm sph}$ and $L_{\rm gal}$ correlate equally well with $M_{\rm BH}$, in terms of intrinsic scatter, only for early-type galaxies; 
once reasonable numbers of spiral galaxies are included, the correlation with $L_{\rm sph}$ is better than that with $L_{\rm gal}$.

@@

with $L_{\rm sph} \propto \sigma^{5-6}$ at the luminous end \citep{schechter1980,malumuthkrishner1981,vonderlinden2007,lauer2007lumell,liu2008}, 

@@

\subsection{Black hole mass -- spheroid stellar mass}
Finally, we present the $M_{\rm BH} - M_{\rm *,sph}$ diagram in Figure \ref{fig:mbhmasssph}, 
and its linear regression analysis in Table \ref{tab:lregmass}. 
The bulges of the early-type galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{1.04 \pm 0.10}$,
consistent with a dry-merging formation scenario,
and define a tight \emph{early-type sequence} with intrinsic scatter $\epsilon_{(Y|X)} = 0.43 \pm 0.06\rm~dex$. 
On the other hand, the bulges of the spiral galaxies trace a steeper \emph{late-type sequence}, 
whose slope is less well constrained due to the smaller size of the subsample and, more importantly, 
to the smaller range in $M_{\rm *,sph}$ that the subsample spans. 
For the bulges of spiral galaxies, the BCES code returns a log-linear relation with a slope $= 3.00 \pm 1.30$, 
while the modified FITEXY routine finds a shallower (but still consistent within the $1\sigma$ uncertainty) 
slope $= 2.28^{+1.67}_{-1.01}$.
The Bayesian estimator of \cite{linmixerr} fails in performing an inverse $(X|Y)$ linear regression for the subsample of spiral galaxies. 
More data would be welcome to better constrain the slope of this \emph{late-type sequence}, 
although we note that direct measurements of black hole masses below $10^6\rm~M_\odot$ are extremely challenging to obtain 
with the current technological resources. 
In this regard, using a sample of $\sim$140 low-redshift ($z \leq 0.35$, with a median redshift $\langle z \rangle = 0.085$) 
bulges hosting Active Galactic Nuclei (AGNs) with virial
black hole masses $10^5 \lesssim M_{\rm BH}/{\rm M_\odot} \lesssim 2 \times 10^6$ \citep{jiang2011a}, 
\cite{grahamscott2015} showed that they roughly follow the quadratic $M_{\rm BH} - M_{\rm *,sph}$ relation defined by their S\'ersic bulges.
The majority of our spiral galaxies host an AGN\footnote{According to the nuclear classification reported on NED 
(NASA Extragalactic Database), among our 17 spiral galaxies, at least 12 host a Seyfert AGN and one hosts a LINER AGN.} and
we anticipate here that the correlation traced by our spiral galaxy bulges 
may track the location of these lower mass AGN in the $M_{\rm BH} - M_{\rm *,sph}$ diagram.
That is, the AGNs appear to be the low-mass continuation of our tentative \emph{late-type sequence} shown in Figure \ref{fig:mbhmasssph} 
and this will be explored with more rigour in a forthcoming paper.

@@

\section{Conclusions}
Using $3.6\rm~\mu m$ \emph{Spitzer} imagery, 
we have performed accurate multicomponent decompositions (i.e.~bulge, disks, bars, spiral arms, rings, halo, nucleus, depleted core, etc.), 
which were checked to be consistent with the two-dimensional galaxy kinematics, 
and we have derived galaxy luminosities, spheroid luminosities and spheroid stellar masses for 66 nearby galaxies 
with a dynamical measurement of their black hole mass. 
Our galaxy sample, besides being the largest ever used to investigate black hole mass scaling relations, 
contains 17 spiral galaxies, half of which have $M_{\rm BH} < 10^7 \rm~M_{\odot}$. 
This constitutes a significant improvement over past studies whose samples were biased towards high-mass, early-type galaxies. \\
Using our state-of-the-art dataset, we have investigated substructure in the $M_{\rm BH} - L_{\rm gal}$, $M_{\rm BH} - L_{\rm sph}$ 
and $M_{\rm BH} - M_{\rm *,sph}$ diagrams. 
Our principal conclusions are: \\

\begin{itemize}
\item The $M_{\rm BH} - M_{\rm *,sph}$ log-relation for the spheroidal components of early-type (elliptical + lenticular) galaxies 
      has a slope of $1.04 \pm 0.10$ and intrinsic scatter $\epsilon_{(Y|X)} = 0.43 \pm 0.06\rm~dex$. 
      We call this tight correlation an \emph{early-type sequence}. 
      The $M_{\rm BH} - M_{\rm *,sph}$ log-relation for the bulges of late-type (spiral) galaxies has a more uncertain slope ($2-3$) 
      due to the smaller size of the subsample and, more importantly, to the smaller range in stellar mass 
      ($3 \times 10^9 \lesssim M_{\rm *,sph}/{\rm M_\odot} \lesssim 3 \times 10^{10}$) that the subsample spans. 
      We refer to this correlation as a ``tentative'' \emph{late-type sequence}. 
      In (gas-poor) early-type galaxies, the black hole and the stellar content of the spheroidal component grow at the same pace, 
      following a linear $M_{\rm BH} - M_{\rm *,sph}$ relation. 
      In (gas-rich) spiral galaxies, the black hole grows faster than its host bulge, 
      following a quadratic/cubic $M_{\rm BH} - M_{\rm *,sph}$ relation. 
      For analogy with the color-magnitude diagram, 
      it would be tempting to refer to our \emph{early-type sequence} as a \emph{red sequence}, 
      although no bimodality in color has been observed in the $M_{\rm BH} - M_{\rm *,sph}$ diagram.
\item The $M_{\rm BH} - M_{\rm *,sph}$ log-relations for core-S\'ersic and S\'ersic spheroids have slopes consistent with each other 
      ($1.19 \pm 0.23$ and $1.48 \pm 0.20$, respectively), 
      thus these two populations do not define two distinct sequences. 
\item In the $M_{\rm BH} - L_{\rm sph}$ (or $M_{\rm BH} - M_{\rm *,sph}$) diagram, for early-type galaxies, 
      we did not observe the change in slope required for consistency with the $M_{\rm BH} - \sigma$ and bent $L_{\rm sph} - \sigma$ correlations.
      This issue of inconsistency remains therefore an open question. 
      It might be that we are still not probing enough low black hole masses ($M_{\rm BH} < 10^7\rm~M_\odot$) 
      for the subsample of early-type galaxies, 
      or that the transition from  $L_{\rm sph} \propto \sigma^2$ at low luminosities to $L_{\rm sph} \propto \sigma^5$ at high luminosities 
      is less sharp than previously thought.
\item Spheroids with S\'ersic index $n_{\rm sph}<2$, argued by some to be pseudo-bulges, 
      are not offset to lower $M_{\rm BH}$ from the $M_{\rm BH} - L_{\rm sph}$ correlation defined by spheroids with $n_{\rm sph}>2$. 
\item The $M_{\rm BH} - L_{\rm gal}$ and $M_{\rm BH} - L_{\rm sph}$ correlations have the same level of intrinsic scatter 
      when considering early-type galaxies only. 
      Once reasonable numbers of spiral galaxies are included, 
      $M_{\rm BH}$ correlates better with $L_{\rm sph}$ than with $L_{\rm gal}$. 
\end{itemize}

@@

\acknowledgments
GS warmly thanks Luca Cortese, Elisabete Lima Da Cunha, Duncan Forbes, Gonzalo Diaz and Pablo Rosado for useful discussion. \\
% referee thank you!!
This research was supported by Australian Research Council funding through grants
DP110103509 and FT110100263.
This work is based on observations made with the IRAC instrument \citep{fazio2004IRAC} 
on-board the Spitzer Space Telescope, 
which is operated by the Jet Propulsion Laboratory, 
California Institute of Technology under a contract with NASA.
This research has made use of the GOLDMine database \citep{goldmine} and the NASA/IPAC Extragalactic Database (NED) 
which is operated by the Jet Propulsion Laboratory, California Institute of Technology, 
under contract with the National Aeronautics and Space Administration. 
The BCES routine \citep{akritasbershady1996} was run via the python module 
written by Rodrigo Nemmen \citep{nemmen2012}, which is available at {\url https://github.com/rsnemmen/BCES}.

>>>>

\begin{abstract}
In our first paper, we performed a detailed (i.e.~bulge, disks, bars, spiral arms, rings, halo, nucleus, depleted core, etc.) 
decomposition of 66 galaxies, with directly measured black hole masses, 
that had been imaged at $3.6\rm~\mu m$ with \emph{Spitzer}.
Our sample is the largest to date and, for the first time, the decompositions were checked for consistency with the galaxy kinematics. 
We present correlations between the black hole mass, $M_{\rm BH}$, 
and the host spheroid (and galaxy) luminosity, $L_{\rm sph}$ (and $L_{\rm gal}$), 
and also stellar mass, $M_{\rm *,sph}$.
Most previous studies have used galaxy samples that were overwhelmingly dominated by high-mass, early-type galaxies.
Instead, our sample includes 17 spiral galaxies, half of which have $M_{\rm BH} < 10^7\rm~M_\odot$, 
and allows us to better investigate the poorly studied low-mass end of the $M_{\rm BH} - M_{\rm *,sph}$ correlation.
The bulges of early-type (E + S0) galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{1.04 \pm 0.10}$, 
consistent with a dry-merging formation scenario,
and define a tight \emph{early-type sequence} with intrinsic scatter $\epsilon_{(Y|X)} = 0.43 \pm 0.06\rm~dex$.
At the low-mass end, the bulges of late-type (Sp) galaxies define a much steeper \emph{late-type sequence}, 
with $M_{\rm BH} \propto M_{\rm *,sph}^{2-3}$, 
indicating that gas-rich processes feed the black hole more efficiently than the host bulge as they coevolve. 
We additionally report that: i) S\'ersic galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{1.48 \pm 0.20}$, a less steep sequence than previously reported; 
ii) bulges with S\'ersic index $n_{\rm sph}<2$, argued by some to be pseudo-bulges, 
are not offset to lower $M_{\rm BH}$ from the correlation defined by bulges with $n_{\rm sph}>2$; 
and iii) $L_{\rm sph}$ and $L_{\rm gal}$ correlate equally well with $M_{\rm BH}$, in terms of intrinsic scatter, only for early-type galaxies; 
once reasonable numbers of spiral galaxies are included, the correlation with $L_{\rm sph}$ is better than that with $L_{\rm gal}$.

@@

with $L_{\rm sph} \propto \sigma^{5-6}$ at the luminous end\footnote{Recent 
work has the $M_{\rm BH} - \sigma$ correlation as steep as $M_{\rm BH} \propto \sigma^{6.5}$ \citep{savorgnangraham2015},
and the high-luminosity end of the $L_{\rm sph} - \sigma$ correlation as steep as $L_{\rm sph} \propto \sigma^{8}$ \citep{monterodorta2015}.} 
\citep{schechter1980,malumuthkrishner1981,vonderlinden2007,lauer2007lumell,liu2008}

@@

\subsection{Black hole mass -- spheroid stellar mass}
Finally, we present the $M_{\rm BH} - M_{\rm *,sph}$ diagram in Figure \ref{fig:mbhmasssph}, 
and its linear regression analysis in Table \ref{tab:lregmass}. 
The bulges of the early-type galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{1.04 \pm 0.10}$,
consistent with a dry-merging formation scenario,
and define a tight \emph{early-type sequence} with intrinsic scatter $\epsilon_{(Y|X)} = 0.43 \pm 0.06\rm~dex$. 
\cite{graham2012bent} reported that the $M_{\rm BH}/M_{\rm dyn,sph}$ ratio for core-S\'ersic galaxies was 0.36\% 
($M_{\rm dyn,sph}$ is the spheroid dynamical mass), 
and discussed the many implications of this.  
Using a larger data sample, \cite{grahamscott2013} reported that the $M_{\rm BH}/M_{\rm *,sph}$ ratio was 0.49\% for core-S\'ersic galaxies.   
Here we find a median $M_{\rm BH}/M_{\rm *,sph}$ ratio of $0.50 \pm 0.04\%$ for the 22 core-S\'ersic galaxies 
and $0.68 \pm 0.04\%$ for the 45 early-type galaxies.  
Among other things, this higher value (previously reported to be $0.1 - 0.2\%$ for all galaxy types, e.g.~\citealt{marconihunt2003}), 
boosts estimates of the black hole mass function and mass density based on galaxy/spheroid luminosity functions. \\
The bulges of the spiral galaxies trace a steeper \emph{late-type sequence}, 
whose slope is less well constrained due to the smaller size of the subsample and, more importantly, 
to the smaller range in $M_{\rm *,sph}$ that the subsample spans. 
For the bulges of spiral galaxies, the BCES code returns a log-linear relation with a slope $= 3.00 \pm 1.30$, 
while the modified FITEXY routine finds a shallower (but still consistent within the $1\sigma$ uncertainty) 
slope $= 2.28^{+1.67}_{-1.01}$.
The Bayesian estimator of \cite{linmixerr} fails in performing an inverse $(X|Y)$ linear regression for the subsample of spiral galaxies. 
More data would be welcome to better constrain the slope of this \emph{late-type sequence}, 
although we note that direct measurements of black hole masses below $10^6\rm~M_\odot$ are extremely challenging to obtain 
with the current technological resources. 
In this regard, using a sample of $\sim$140 low-redshift ($z \leq 0.35$, with a median redshift $\langle z \rangle = 0.085$) 
bulges hosting Active Galactic Nuclei (AGNs) with virial
black hole masses $10^5 \lesssim M_{\rm BH}/{\rm M_\odot} \lesssim 2 \times 10^6$ \citep{jiang2011a}, 
\cite{grahamscott2015} showed that they roughly follow the quadratic $M_{\rm BH} - M_{\rm *,sph}$ relation defined by their S\'ersic bulges.
The majority of our spiral galaxies host an AGN\footnote{According to the nuclear classification reported on NED 
(NASA Extragalactic Database), among our 17 spiral galaxies, at least 12 host a Seyfert AGN and one hosts a LINER AGN.} and
we anticipate here that the correlation traced by our spiral galaxy bulges 
may track the location of these lower mass AGN in the $M_{\rm BH} - M_{\rm *,sph}$ diagram.
That is, the AGNs appear to be the low-mass continuation of our tentative \emph{late-type sequence} shown in Figure \ref{fig:mbhmasssph} 
and this will be explored with more rigour in a forthcoming paper.

@@

\section{Conclusions}
Using $3.6\rm~\mu m$ \emph{Spitzer} images, 
we have performed accurate multicomponent decompositions (i.e.~bulge, disks, bars, spiral arms, rings, halo, nucleus, depleted core, etc.), 
which were checked to be consistent with the two-dimensional galaxy kinematics, 
for 66 nearby galaxies with a dynamical measurement of their black hole mass.
We have derived galaxy luminosities, spheroid luminosities and spheroid stellar masses. 
Our galaxy sample, besides being to date the largest sample with reliable bulge masses used to investigate black hole mass scaling relations, 
contains 17 spiral galaxies, half of which have $M_{\rm BH} < 10^7 \rm~M_{\odot}$. 
This constitutes a significant improvement over past studies whose samples were biased towards high-mass, early-type galaxies. \\
Using our state-of-the-art dataset, we have investigated substructure in the $M_{\rm BH} - L_{\rm gal}$, $M_{\rm BH} - L_{\rm sph}$ 
and $M_{\rm BH} - M_{\rm *,sph}$ diagrams. 
Our principal conclusions are: \\

\begin{itemize}
\item The $M_{\rm BH} - M_{\rm *,sph}$ log-relation for the spheroidal components of early-type (elliptical + lenticular) galaxies 
      has a slope of $1.04 \pm 0.10$ and intrinsic scatter $\epsilon_{(Y|X)} = 0.43 \pm 0.06\rm~dex$. 
      We call this tight correlation an \emph{early-type sequence}. 
      The $M_{\rm BH} - M_{\rm *,sph}$ log-relation for the bulges of late-type (spiral) galaxies has a slope of $2-3$,
      which is less well constrained due to
      the smaller size of the subsample and, more importantly, the smaller range in spheroid stellar mass 
      ($3 \times 10^9 \lesssim M_{\rm *,sph}/{\rm M_\odot} \lesssim 3 \times 10^{10}$) that the subsample spans. 
      We refer to this correlation as a ``tentative'' \emph{late-type sequence}. 
      In (gas-poor) early-type galaxies, the black hole and the stellar content of the spheroidal component grow at the same pace, 
      following a linear $M_{\rm BH} - M_{\rm *,sph}$ relation. 
      In (gas-rich) spiral galaxies, the black hole grows faster than its host bulge, 
      following a quadratic/cubic $M_{\rm BH} - M_{\rm *,sph}$ relation. 
      For analogy with the color-magnitude diagram, 
      it may be tempting to refer to our \emph{early-type sequence} as a \emph{red sequence} 
      and to our \emph{late-type sequence} as a \emph{blue sequence}.
      Future work will explore the color of spheroids and galaxies in the $M_{\rm BH} - M_{\rm *,sph}$ diagram.  
\item The $M_{\rm BH}/M_{\rm *,sph}$ ratio for the early-type galaxies is $0.68 \pm 0.04\%$.

was 0.49\% for core-S\'ersic galaxies.   
Here we find a median $M_{\rm BH}/M_{\rm *,sph}$ ratio of $0.50 \pm 0.04\%$ for the 22 core-S\'ersic galaxies 
and $0.68 \pm 0.04\%$ for the 45 early-type galaxies.  
Among other things, this higher value (previously reported to be $0.1 - 0.2\%$ for all galaxy types, e.g.~\citealt{marconihunt2003}),

\item The $M_{\rm BH} - M_{\rm *,sph}$ log-relations for core-S\'ersic and S\'ersic spheroids have slopes consistent with each other 
      ($1.19 \pm 0.23$ and $1.48 \pm 0.20$, respectively), 
      thus these two populations do not appear to define two distinct sequences. 
\item In the $M_{\rm BH} - L_{\rm sph}$ (or $M_{\rm BH} - M_{\rm *,sph}$) diagram, for early-type galaxies, 
      we did not observe the change in slope required for consistency with the log-linear $M_{\rm BH} - \sigma$ and bent $L_{\rm sph} - \sigma$ correlations.
      This issue of inconsistency remains therefore an open question. 
      It might be that we are still not probing enough low-mass black holes ($M_{\rm BH} < 10^7\rm~M_\odot$) 
      for the subsample of early-type galaxies, 
      or that the transition from  $L_{\rm sph} \propto \sigma^2$ at low luminosities to $L_{\rm sph} \propto \sigma^{(5-6)}$ at high luminosities 
      is less sharp than previously thought.
      We intend to investigate this point in our future work. 
\item Spheroids with S\'ersic index $n_{\rm sph}<2$, argued by some to be pseudo-bulges, 
      are not offset to lower $M_{\rm BH}$ from the $M_{\rm BH} - L_{\rm sph}$ correlation defined by spheroids with $n_{\rm sph}>2$. 
\item The $M_{\rm BH} - L_{\rm gal}$ and $M_{\rm BH} - L_{\rm sph}$ correlations have the same level of intrinsic scatter 
      when considering early-type galaxies only. 
      Once reasonable numbers of spiral galaxies are included, 
      $M_{\rm BH}$ correlates better with $L_{\rm sph}$ than with $L_{\rm gal}$. 
\end{itemize}

@@

\acknowledgments
GS warmly thanks Luca Cortese, Elisabete Lima Da Cunha, Duncan Forbes and Gonzalo Diaz for useful discussion. \\
% referee thank you!!
This research was supported by Australian Research Council funding through grants
DP110103509 and FT110100263.
This work is based on observations made with the IRAC instrument \citep{fazio2004IRAC} 
on-board the Spitzer Space Telescope, 
which is operated by the Jet Propulsion Laboratory, 
California Institute of Technology under a contract with NASA.
This research has made use of the GOLDMine database \citep{goldmine} and the NASA/IPAC Extragalactic Database (NED) 
which is operated by the Jet Propulsion Laboratory, California Institute of Technology, 
under contract with the National Aeronautics and Space Administration. 
The BCES routine \citep{akritasbershady1996} was run via the python module 
written by Rodrigo Nemmen \citep{nemmen2012}, which is available at {\url https://github.com/rsnemmen/BCES}.
